---
title: "1-human-response"
format:
    html:
      toc-title: Table of contents
      toc: true
      toc-depth: 2
      code-fold: true # allow code folding
      number-sections: true
      highlight-style: github
      toc-location: body
      cap-location: top
      page-layout: full
      embed-resources: true
      self-contained-math: true
      toc-expand: true
      theme: cosmo
execute:
  echo: false        # show R code
  warning: false    # hide warnings
  message: false    # hide messages
  eval: true        # run the code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F,message = F)
options(download.file.method = "libcurl")

if (!require("groundhog")) install.packages("groundhog")
library("groundhog")
pkgs <- c("tidyverse","car","psych","lme4","table1","jtools","sjPlot","sjmisc","ggpubr","rstatix","lmtest","robustbase","see","rempsyc","sandwich","stargazer","boot","patchwork","data.table","irr")
groundhog.library(pkgs, "2025-08-11")

options(scipen = 999, digits = 3)
set.seed(111)
mycolor=c("gray","#756bb1", "#fdae6b")
```

```{r}
wide<-read.csv( "data/human-wide.csv")
long<-read.csv( "data/human-long.csv")
df<-long %>% filter(strength_value>0)
```

# descriptive table

## demographics

```{r}
#df<-read.csv( "data/df_all_long_filtered_last.csv") 
#table1(~ gender + as.numeric(age)+race,data=filter(df,label_type=="human") %>% distinct(human_id,.keep_all = T))

table1(~ gender + as.numeric(age)+race,data=distinct(wide,ResponseId,.keep_all = T))
```


# Descriptive plots & analysis 

## Gender
### Mean Emotion Strength by Emotion Type

```{r,fig.height=6,fig.width=8}
# Count number of occurrences of each emotion
emotion_count <- df %>%
  group_by(emo_type, emotion, actor_gender) %>%
  summarise(count = n(), .groups = "drop")

# Compute chi-square test for each emo_type
gender_stats <- df %>%
  group_by(emo_type) %>%
  summarise(
    test = list(chisq.test(table(emotion, actor_gender))),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      res <- .
      paste0("χ²(", res$parameter, ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )

# Determine y position for labels (above bars)
label_positions <- emotion_count %>%
  group_by(emo_type) %>%
  summarise(y_pos = max(count) + 1, .groups = "drop") %>%
  left_join(gender_stats %>% select(emo_type, label), by = "emo_type") %>%
  mutate(x = 4,y=620)  # approximate center above bars

# Bar plot with male/female side by side
ggplot(emotion_count, aes(x = reorder(emotion, -count), y = count, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.7)) +
  facet_wrap(~emo_type, scales = "free_y") +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0.5, hjust = 0.5) +
  labs(
    x = "Emotion",
    y = "Count",
    title = "Count of Each Emotion by Emotion Type and Actor Gender"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Pastel1")+
 coord_cartesian(ylim = c(0, 650))   # consistent y-axis

```


```{r}
emotion_summary <- df %>%
  group_by(emo_type, emotion, actor_gender) %>%
  summarise(
    mean_strength = mean(strength_value, na.rm = TRUE),
    se_strength = sd(strength_value, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

# t-tests for strength_value by gender for each emotion × emo_type
t_results <- df %>%
  group_by(emo_type, emotion) %>%
  summarise(
    p_value = tryCatch(t.test(strength_value ~ actor_gender)$p.value,
                       error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(sig = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    TRUE            ~ ""
  )) 

# Position for stars
# Compute label positions for significance stars
label_positions <- emotion_summary %>%
  left_join(t_results %>%
  filter(sig != "") %>% 
  select(emo_type, emotion, sig), by = c("emo_type", "emotion")) %>%
  filter(!is.na(sig)) %>%                       # keep only significant tests
  group_by(emo_type, emotion) %>%
  summarise(
    y_pos = max(mean_strength + se_strength) + 0.2,  # place above the tallest bar
    sig = first(sig),                                # take the sig for this group
    .groups = "drop"
  )


# Plot
ggplot(emotion_summary, aes(x = emotion, y = mean_strength, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.7)) +
  geom_errorbar(aes(ymin = mean_strength - se_strength,
                    ymax = mean_strength + se_strength),
                width = 0.2, position = position_dodge(width = 0.7)) +
  geom_text(data = label_positions,
            aes(x = emotion, y = y_pos, label = sig),
            inherit.aes = FALSE, vjust = 0, hjust = 0.5, size = 5) +
  facet_wrap(~emo_type, scales = "free_y") +
  labs(
    x = "Emotion",
    y = "Mean Strength",
    title = "Mean Emotion Strength by Emotion Type and Actor Gender"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Pastel1") +
  coord_cartesian(ylim = c(0,5))

t_results
```

### Action by Actor Gender

```{r, echo=F,include=F}
wide %>%
  select(
    id,
    actor_gender,
    stranger_action = stran_act,
    stranger_action_should = stran_act_should,
    acquaintance_action = acq_act,
    acquaintance_action_should = acq_act_should,
    friend_action = friend_act,
    friend_action_should = friend_act_should
  ) %>%
  pivot_longer(
    cols = stranger_action:friend_action_should,
    names_to = "variable",
    values_to = "value"
  ) %>% 
  distinct(id, actor_gender, variable, value) %>% 
  mutate(value = factor(value)) %>%
  # reorder by count within each facet
  group_by(variable, actor_gender) %>%
  mutate(value = fct_infreq(value)) %>%  # orders factor by descending frequency
  ungroup() %>%
  ggplot(aes(x = value,fill=value)) +
  geom_bar() +
  facet_grid(variable ~ actor_gender, scales = "free_y") +
  labs(title = "Distribution of Actions and Shoulds",
       x = "",
       y = "Count") +
  theme_classic()+
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())+
  scale_fill_brewer(palette = "Pastel1")
```


```{r,fig.height=5,fig.width=8}
library(tidyverse)
library(forcats)

# Prepare plot data
plot_data <- wide %>%
  select(
    id,
    actor_gender,
    would_stranger = stran_act,
    should_stranger = stran_act_should,
    would_acquaintance = acq_act,
    should_acquaintance = acq_act_should,
    would_friend = friend_act,
    should_friend = friend_act_should
  ) %>%
  pivot_longer(
    cols = would_stranger:should_friend,
    names_to = "variable",
    values_to = "value"
  )%>% 
  distinct(id, actor_gender, variable, value) %>% 
  mutate(value = factor(value)) %>%
  group_by(variable, actor_gender) %>%
  mutate(value = fct_infreq(value)) %>%
  ungroup()

# Chi-square test results
test_results <- plot_data %>%
  group_by(variable) %>%
  summarise(
    test = list(chisq.test(table(value, actor_gender))),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      res <- .
      paste0("χ²(", res$parameter, ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )


# Add y-position for label (centered above each facet)
label_positions <- plot_data %>%
  group_by(variable) %>%
  summarise(y_pos = max(table(value)) + 1, .groups = "drop") %>%  # or a fixed number
  left_join(test_results %>% select(variable, label), by = "variable") %>%
  mutate(x = 4.5,y=650)  # center x (approximate middle of the categories)

# Plot
ggplot(plot_data, aes(x = value, fill = actor_gender)) +
  geom_bar(position = position_dodge(width = 0.7)) +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0.5, hjust = 0.5) +
  labs(title = "Distribution of Actions and Shoulds by Gender",
       x = NULL,
       y = "Count") +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # rotate 45 degrees
    axis.ticks.x = element_blank()
  )+
   coord_cartesian(ylim = c(0, 700))   # consistent y-axis
```

### Wrongness vs. Appropriateness Ratings by Actor Gender

```{r}
# Summarize mean and SE
summary_stats <- wide %>%
  pivot_longer(cols = c(appropriate, wrong),
               names_to = "judgment", values_to = "rating") %>%
  group_by(actor_gender, judgment) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    se_rating = sd(rating, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Compute t-test results for each judgment
test_results <- wide %>%
  pivot_longer(cols = c(appropriate, wrong),
               names_to = "judgment", values_to = "rating") %>%
  group_by(judgment) %>%
  summarise(
    t_test = list(t.test(rating ~ actor_gender)),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(t_test, ~{
      res <- .
      paste0("t(", round(res$parameter,1), ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )

# Determine y position for the label (above bars)
label_positions <- summary_stats %>%
  group_by(judgment) %>%
  summarise(y_pos = max(mean_rating + se_rating) + 0.2, .groups = "drop") %>%
  left_join(test_results %>% select(judgment, label), by = "judgment")

# Bar plot with error bars and test results
ggplot(summary_stats, aes(x = judgment, y = mean_rating, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = mean_rating - se_rating,
                    ymax = mean_rating + se_rating),
                width = 0.1,
                position = position_dodge(width = 0.9)) +
  geom_text(data = label_positions,
            aes(x = judgment, y = y_pos, label = label),
            inherit.aes = FALSE,
            vjust = 0) +
  labs(
    x = "Judgment",
    y = "Mean Rating",
    title = "Mean Ratings of Wrongness and Appropriateness by Actor Gender"
  ) +
  theme_classic() +
  scale_y_continuous(limits = c(0, 4))+
  scale_fill_brewer(palette = "Pastel1")
```

## Social distance

### emotion

### action
```{r,fig.height=5,fig.width=8}
# 1 Prepare plot data: merge gender
plot_data <- wide %>%
  select(
    id = ResponseId,
    would_stranger = stran_act,
    should_stranger = stran_act_should,
    would_acquaintance = acq_act,
    should_acquaintance = acq_act_should,
    would_friend = friend_act,
    should_friend = friend_act_should
  ) %>%
  pivot_longer(
    cols = -id,
    names_to = "variable",
    values_to = "value"
  ) %>%
  distinct(id, variable, value) %>%
  mutate(
    value = factor(value),
    measure = if_else(str_detect(variable, "would"), "would", "should"),
    tie_type = case_when(
      str_detect(variable, "friend") ~ "friend",
      str_detect(variable, "acquaintance") ~ "acquaintance",
      str_detect(variable, "stranger") ~ "stranger"
    ),
    tie_type = factor(tie_type, levels = c("friend", "acquaintance", "stranger"))
  )

# 2 Count occurrences for plotting
counts <- plot_data %>%
  group_by(measure, tie_type, value) %>%
  summarise(n = n(), .groups = "drop")

# Global chi-square tests (one per measure)
global_chi <- plot_data %>%
  group_by(measure) %>%
  summarise(
    test = list({
      tbl <- table(value, tie_type)
      if (all(rowSums(tbl) > 0) && all(colSums(tbl) > 0)) chisq.test(tbl) else NULL
    }),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      if (is.null(.x)) return("Insufficient data")
      paste0("χ²(", .x$parameter, ") = ",
             round(.x$statistic, 2),
             ", p < .001")
    })
  )

# Show results
global_chi 

# 3 Pairwise chi-square comparisons among tie types (within each measure)
pairwise_tests0 <- map_dfr(unique(plot_data$measure), function(m) {
  data_m <- filter(plot_data, measure == m)

  # Only use tie types that actually appear in this measure
  tie_types_m <- unique(data_m$tie_type)

  # Generate all pairwise combinations
  pairs <- combn(tie_types_m, 2, simplify = FALSE)

  map_dfr(pairs, function(pair) {
    sub <- filter(data_m, tie_type %in% pair)
    tbl <- table(sub$value, sub$tie_type)

    # remove rows or columns that are entirely zero
    tbl2 <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]

    if (nrow(tbl2) < 2 || ncol(tbl2) < 2) {
      # skip invalid comparisons
      return(tibble(
        measure = m,
        tie1 = pair[1],
        tie2 = pair[2],
        p_value = NA_real_,
        sig = ""
      ))
    }

    # Run test (safe)
    p <- tryCatch({
      suppressWarnings(chisq.test(tbl2))$p.value
    }, error = function(e) {
      tryCatch({
        fisher.test(tbl2, simulate.p.value = TRUE)$p.value
      }, error = function(e2) NA_real_)
    })

    tibble(
      measure = m,
      tie1 = pair[1],
      tie2 = pair[2],
      p_value = p,
      sig = case_when(
        !is.na(p) & p < 0.001 ~ "***",
        !is.na(p) & p < 0.01  ~ "**",
        !is.na(p) & p < 0.05  ~ "*",
        TRUE ~ ""
      )
    )
  })
})

pairwise_tests0

# Determine x-positions for significance stars
label_positions <- plot_data %>%
  group_by(measure) %>%
  summarise(y_pos = max(table(value)) + 1, .groups = "drop") %>%  # or a fixed number
  left_join(global_chi %>% select(measure, label), by = "measure") %>%
  mutate(x = 4.5,y=650)  # center x (approximate middle of the categories)

# Plot
ggplot(counts, aes(x = value, y = n, fill = tie_type)) +
  geom_col(position = position_dodge(width = 0.8)) +
   geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0) +
  facet_wrap(~measure) +
  labs(
    title = "Distribution of Actions by Social Distance",
    x = "Social Distance (Tie Type)",
    y = "Count"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, 750))
```

### emotion 
```{r}
# --- Prepare plot data: count of each emotion by tie type ---
plot_data <- df %>%
  filter(emo_type!="emotion_actor") %>% 
  select(
    id,
    actor_gender,  # optional, can ignore
    emo_type,
    emotion
  )  %>%
  mutate(
    tie_type = case_when(
      emo_type == "emotion_friend" ~ "Friend",
      emo_type == "emotion_acquaintance" ~ "Acquaintance",
      emo_type == "emotion_stranger" ~ "Stranger"
    ),
    tie_type = factor(tie_type, levels = c("Friend", "Acquaintance", "Stranger")),
    emotion = factor(emotion)
  )

# --- Global chi-square test: emotion counts by tie type ---
chi_test <- chisq.test(table(plot_data$emotion, plot_data$tie_type))

chi_label <- paste0(
  "χ²(", chi_test$parameter, ") = ", round(chi_test$statistic, 2),
  ", p < .001"
)

# --- Count occurrences for plotting ---
counts <- plot_data %>%
  group_by(tie_type, emotion) %>%
  summarise(n = n(), .groups = "drop")

# --- Plot ---
ggplot(counts, aes(x = emotion, y = n, fill = tie_type)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Distribution of Emotions by Tie Type",
    subtitle = chi_label,
    x = NULL,
    y = "Count"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, max(counts$n) + 50))

```


# IRR

## emotion
```{r,eval=F}
# Krippendorff’s α requires that pairs of raters share overlapping items so that you can estimate their pairwise agreement structure. Every “pair of raters” contributes disagreement terms based on how similarly they rate the same items. If each rater only overlaps with one or two others on a few items, you don’t have enough pairwise co-ratings to estimate (observed disagreement) reliably. In the extreme case (each rater rates unique items), pairwise matrix is undefined.

# Function to compute Krippendorff's alpha for one emotion
compute_alpha_emotion <- function(df, emotion_name,emo_type_name) {
  tmp <- df %>%
    filter(emo_type == emo_type_name) %>% 
    filter(emotion == emotion_name) %>%
    select(behavior, id, strength_value) %>%
  #  mutate(strength_value = ifelse(strength_value == 0, NA, strength_value)) %>% 
    group_by(behavior) %>%
    summarise(ratings = list(strength_value), .groups = "drop")
  
  # Combine all item ratings into a matrix: rows = items, columns = ratings for that item
  max_raters <- max(sapply(tmp$ratings, length))
  
  mat <- t(sapply(tmp$ratings, function(x) {
    # pad with NA if fewer ratings
    c(x, rep(NA, max_raters - length(x)))
  }))
  
  # compute Krippendorff's alpha
  kripp.alpha(mat, method = "ordinal")$value
}

# Loop over all emo_types and emotions
emo_types <- unique(long$emo_type)
emotions <- unique(long$emotion)

results_df <- expand.grid(emotion = emotions, emo_type = emo_types)

# Use pmap to pass each row as scalar arguments
results_df$alpha <- pmap_dbl(
  list(results_df$emotion, results_df$emo_type),
  function(e, etype) compute_alpha_emotion(long, e, etype)
)

# Pivot to wide: rows = norm violator type, columns = emotions
results_wide <- results_df %>%
  pivot_wider(names_from = emotion, values_from = alpha) %>%
  arrange(factor(emo_type, levels = emo_types))

results_wide
```

```{r}
# Compute standard deviation per emotion × behavior
item_emotion_strength<-long %>%
  group_by(emo_type,emotion, behavior) %>%
  summarise(item_mean=mean(strength_value,na.rm=TRUE),
    item_sd = sd(strength_value, na.rm = TRUE)) 

item_emotion_strength
```

# ICC

```{r}
df_all<-read.csv( "data/df_all_long_updated.csv") %>% 
  mutate(rater=ifelse(label_type=="human","human","llm")) %>% 
  mutate(unique_id=paste0(label_type,"_",human_id)) %>% 
  mutate(ID=as.factor(ID)) %>% 
  filter(rater == "human")
```

For continuous behavior follow linear distribution

$$
ICC_{\text{behavior}} = \frac{\sigma^2_{\text{behavior}}}{\sigma^2_{\text{behavior}} + \sigma^2_{\text{rater}} + \sigma^2_{\text{residual}}}
$$

For binary behavior follow binomial distribution

$$
ICC_{\text{behavior}} = \frac{\sigma^2_{\text{behavior}}}{\sigma^2_{\text{behavior}} + \sigma^2_{\text{rater}} + \pi^2/3}
$$

```{r,eval=F}
#ICC = 0.05 → Only 5% of rating variance is due to systematic differences between behaviors; raters largely disagree or interpret scales differently.
icc_emotion <- long %>%
  # sensitivity test for binary coding
  mutate(strength_value_bi=ifelse(strength_value==0,0,1)) %>% 
  group_by(emotion, emo_type) %>%
  group_modify(~{
    data_sub <- .x
    # Fit random-effects modelß
    m <- tryCatch(
      #lmer(strength_value ~ 1 + (1 | behavior) + (1 | id), data = data_sub),
      # binary outcome
      glmer(strength_value_bi ~ 1 + (1|behavior) + (1|id), data = data_sub, family = binomial),
      error = function(e) NULL
    )

    # Extract variance components for linear model
    v <- as.data.frame(VarCorr(m))
    sigma_b <- v[v$grp == "behavior", "vcov"]
    sigma_r <- v[v$grp == "id", "vcov"]
    # linear residual variance
    #sigma_e <- attr(VarCorr(m), "sc")^2
    # logistic residual variance
    sigma_e <- (pi^2) / 3  

    # Compute ICC for behavior
    ICC <- sigma_b / (sigma_b + sigma_r + sigma_e)
    tibble(ICC = ICC)
  }) %>%
  ungroup()

icc_emotion %>% 
  pivot_wider(names_from = emotion, values_from = ICC) %>%
  nice_table(.)

# plot 
ggplot(icc_emotion, aes(x = emotion, y = ICC, fill = emo_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_hline(yintercept = c(0.2, 0.5, 0.75), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = Inf, y = 0.2, label = "Fair", hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.5, label = "Good", hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.75, label = "Excellent", hjust = 1.1, vjust = -0.2, color = "gray50") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Inter-rater Reliability by Emotion and Agent Type",
       x = "Emotion",
       y = "Intra-class Correlation (ICC)",
       fill = "Agent Type") +
  theme_classic(base_size = 14)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r using all}
#ICC = 0.05 → Only 5% of rating variance is due to systematic differences between behaviors; raters largely disagree or interpret scales differently.
icc_emotion <- df_all %>%
  # binary sensitivity coding
  mutate(strength_value_bi = ifelse(strength_value == 0, 0, 1)) %>%
  group_by(emotion, emotion_type) %>%
  group_modify(~ {
    data_sub <- .x

    m <- tryCatch(
      glmer(
        strength_value_bi ~ 1 +
          (1 | ID) +            # behavior-level variance (vignette ID)
          (1 | unique_id),      # rater-level variance
        data = data_sub,
        family = binomial
      ),
      error = function(e) NULL
    )

    if (is.null(m)) {
      return(tibble(ICC = NA_real_))
    }

    v <- as.data.frame(VarCorr(m))

    sigma_b <- v[v$grp == "ID", "vcov"]
    sigma_r <- v[v$grp == "unique_id", "vcov"]

    # logistic residual variance
    sigma_e <- (pi^2) / 3

    ICC <- sigma_b / (sigma_b + sigma_r + sigma_e)

    tibble(ICC = ICC)
  }) %>%
  ungroup()

icc_emotion %>%
  pivot_wider(names_from = emotion, values_from = ICC) %>%
  nice_table(.)

ggplot(icc_emotion, aes(x = emotion, y = ICC, fill = emotion_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_hline(yintercept = c(0.2, 0.5, 0.75),
             linetype = "dashed", color = "gray50") +
  annotate("text", x = Inf, y = 0.2, label = "Fair",
           hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.5, label = "Good",
           hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.75, label = "Excellent",
           hjust = 1.1, vjust = -0.2, color = "gray50") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Inter-rater Reliability by Emotion and Agent Type",
    x = "Emotion",
    y = "Intra-class Correlation (ICC)",
    fill = "Emotion Type"
  ) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## action - would/should

```{r, eval=F}
action_vars <- c("stran_act", "acq_act", "friend_act")
action_vars <- c("stran_act_should", "acq_act_should", "friend_act_should")

# Reshape to long and create binary columns
act_long <- wide %>%
  select(id, behavior, all_of(action_vars)) %>%  # keep necessary columns
  pivot_longer(cols = all_of(action_vars), 
               names_to = "actor_type", 
               values_to = "action") %>% 
  # create 0/1 for each category
  mutate(action_value = 1) %>% 
  pivot_wider(names_from = action, 
              values_from = action_value, 
              values_fill = 0) 

# ICC
icc_act <- act_long %>%
  pivot_longer(cols = -c(id, behavior, actor_type),
               names_to = "category",
               values_to = "value") %>%
  group_by(actor_type, category) %>%
  group_modify(~{
    data_sub <- .x
    # Fit random-effects model
    m <- tryCatch(
      glmer(value ~ 1 + (1|behavior) + (1|id), data = data_sub, family = binomial),
      error = function(e) NULL
    )

    # Extract variance components
    v <- as.data.frame(VarCorr(m))
    sigma_b <- v[v$grp == "behavior", "vcov"]
    sigma_r <- v[v$grp == "id", "vcov"]
    # For binary outcomes, residual variance on logit scale is pi^2 / 3
    sigma_e <- (pi^2)/3

    # Compute ICC for "behavior"
    ICC <- sigma_b / (sigma_b + sigma_r + sigma_e)

    tibble(ICC = ICC)
  }) %>%
  ungroup()

icc_act %>% 
  pivot_wider(names_from = category, values_from = ICC) %>%
  nice_table(.)

```

```{r use all data}
action_vars <- c(
  "strangers_reaction",
  "acquaintances_reaction",
  "friends_reaction"
)

act_long <- all %>%
  select(unique_id, ID, all_of(action_vars)) %>%
  
  pivot_longer(
    cols = all_of(action_vars),
    names_to = "actor_type_raw",
    values_to = "action"
  ) %>%
  distinct() %>% 
  mutate(
    actor_type = case_when(
      actor_type_raw == "strangers_reaction"     ~ "stranger",
      actor_type_raw == "acquaintances_reaction" ~ "acquaintance",
      actor_type_raw == "friends_reaction"       ~ "friend"
    ),
    action_value = 1
  ) %>%
  select(-actor_type_raw) %>%
  pivot_wider(
    names_from = action,
    values_from = action_value,
    values_fill = 0
  )

icc_act <- act_long %>%
  pivot_longer(
    cols = -c(unique_id, ID, actor_type),
    names_to = "category",
    values_to = "value"
  ) %>%
  group_by(actor_type, category) %>%
  group_modify(~ {
    data_sub <- .x

    m <- tryCatch(
      glmer(
        value ~ 1 +
          (1 | ID) +            # vignette-level variance
          (1 | unique_id),      # rater-level variance
        data = data_sub,
        family = binomial
      ),
      error = function(e) NULL
    )

    if (is.null(m)) {
      return(tibble(ICC = NA_real_))
    }

    v <- as.data.frame(VarCorr(m))

    sigma_b <- v[v$grp == "ID", "vcov"]
    sigma_r <- v[v$grp == "unique_id", "vcov"]

    # residual variance for logit link
    sigma_e <- (pi^2) / 3

    ICC <- sigma_b / (sigma_b + sigma_r + sigma_e)

    tibble(ICC = ICC)
  }) %>%
  ungroup()


icc_act %>%
  pivot_wider(names_from = category, values_from = ICC) %>%
  nice_table(.)

# plot 
ggplot(icc_act, aes(x = category, y = ICC, fill = actor_type)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  geom_hline(yintercept = c(0.2, 0.5, 0.75), 
             linetype = "dashed", color = "gray50") +
  annotate("text", x = Inf, y = 0.2, label = "Fair", hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.5, label = "Good", hjust = 1.1, vjust = -0.2, color = "gray50") +
  annotate("text", x = Inf, y = 0.75, label = "Excellent", hjust = 1.1, vjust = -0.2, color = "gray50") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Inter-rater Reliability by Action and Observer",
       x = "Action",
       y = "Intra-class Correlation (ICC)",
       fill = "Observer") +
  theme_classic(base_size = 14)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

```{r}
# actor types 
count(act_long, actor_type)

# base rates per category (detect degenerate ICCs)
act_long %>%
  pivot_longer(
    cols = -c(unique_id, ID, actor_type),
    names_to = "category",
    values_to = "value"
  ) %>%
  group_by(actor_type, category) %>%
  summarise(p = mean(value), .groups = "drop")

```



