---
title: "human-analysis"
format:
  html:
    toc: true       # include table of contents
    code-fold: true # allow code folding
    theme: cosmo
execute:
  echo: false        # show R code
  warning: false    # hide warnings
  message: false    # hide messages
  eval: true        # run the code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F,message = F)
options(download.file.method = "libcurl")

if (!require("groundhog")) install.packages("groundhog")
library("groundhog")
pkgs <- c("tidyverse","car","psych","lme4","table1","jtools","sjPlot","sjmisc","ggpubr","rstatix","lmtest","robustbase","see","rempsyc","sandwich","stargazer","boot","patchwork","data.table","irr")
groundhog.library(pkgs, "2025-08-11")

options(scipen = 999, digits = 3)
set.seed(111)
mycolor=c("gray","#756bb1", "#fdae6b")
```

```{r}
wide<-read.csv( "data/human-wide.csv")
df<-read.csv( "data/human-long.csv")
```

# descriptive table

## demographics

```{r}
table1(~ gender + as.numeric(age)+race,data=distinct(df,id,.keep_all = TRUE))
```

## Descriptive stat table (TBD)

```{r}
#table1(~ emotion | actor_gender+emo_type,df)
#table1(~ strength_value |emotion+actor_gender,filter(df, emo_type == "emotion_actor"))
```

# Descriptive plots & analysis 

## GENDER
### Mean Emotion Strength by Emotion Type

```{r,fig.height=6,fig.width=8}
# Count number of occurrences of each emotion
emotion_count <- df %>%
  group_by(emo_type, emotion, actor_gender) %>%
  summarise(count = n(), .groups = "drop")

# Compute chi-square test for each emo_type
gender_stats <- df %>%
  group_by(emo_type) %>%
  summarise(
    test = list(chisq.test(table(emotion, actor_gender))),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      res <- .
      paste0("χ²(", res$parameter, ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )

# Determine y position for labels (above bars)
label_positions <- emotion_count %>%
  group_by(emo_type) %>%
  summarise(y_pos = max(count) + 1, .groups = "drop") %>%
  left_join(gender_stats %>% select(emo_type, label), by = "emo_type") %>%
  mutate(x = 4,y=620)  # approximate center above bars

# Bar plot with male/female side by side
ggplot(emotion_count, aes(x = reorder(emotion, -count), y = count, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.7)) +
  facet_wrap(~emo_type, scales = "free_y") +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0.5, hjust = 0.5) +
  labs(
    x = "Emotion",
    y = "Count",
    title = "Count of Each Emotion by Emotion Type and Actor Gender"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Pastel1")+
 coord_cartesian(ylim = c(0, 650))   # consistent y-axis


```


```{r}
emotion_summary <- df %>%
  group_by(emo_type, emotion, actor_gender) %>%
  summarise(
    mean_strength = mean(strength_value, na.rm = TRUE),
    se_strength = sd(strength_value, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  )

# t-tests for strength_value by gender for each emotion × emo_type
t_results <- df %>%
  group_by(emo_type, emotion) %>%
  summarise(
    p_value = tryCatch(t.test(strength_value ~ actor_gender)$p.value,
                       error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(sig = case_when(
    p_value < 0.001 ~ "***",
    p_value < 0.01  ~ "**",
    p_value < 0.05  ~ "*",
    TRUE            ~ ""
  )) 

# Position for stars
# Compute label positions for significance stars
label_positions <- emotion_summary %>%
  left_join(t_results %>%
  filter(sig != "") %>% 
  select(emo_type, emotion, sig), by = c("emo_type", "emotion")) %>%
  filter(!is.na(sig)) %>%                       # keep only significant tests
  group_by(emo_type, emotion) %>%
  summarise(
    y_pos = max(mean_strength + se_strength) + 0.2,  # place above the tallest bar
    sig = first(sig),                                # take the sig for this group
    .groups = "drop"
  )


# Plot
ggplot(emotion_summary, aes(x = emotion, y = mean_strength, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.7)) +
  geom_errorbar(aes(ymin = mean_strength - se_strength,
                    ymax = mean_strength + se_strength),
                width = 0.2, position = position_dodge(width = 0.7)) +
  geom_text(data = label_positions,
            aes(x = emotion, y = y_pos, label = sig),
            inherit.aes = FALSE, vjust = 0, hjust = 0.5, size = 5) +
  facet_wrap(~emo_type, scales = "free_y") +
  labs(
    x = "Emotion",
    y = "Mean Strength",
    title = "Mean Emotion Strength by Emotion Type and Actor Gender"
  ) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Pastel1") +
  coord_cartesian(ylim = c(0,5))

t_results
```

### Action by Actor Gender

```{r, echo=F,include=F}
wide %>%
  select(
    id,
    actor_gender,
    stranger_action = stran_act,
    stranger_action_should = stran_act_should,
    acquaintance_action = acq_act,
    acquaintance_action_should = acq_act_should,
    friend_action = friend_act,
    friend_action_should = friend_act_should
  ) %>%
  pivot_longer(
    cols = stranger_action:friend_action_should,
    names_to = "variable",
    values_to = "value"
  ) %>% 
  distinct(id, actor_gender, variable, value) %>% 
  mutate(value = factor(value)) %>%
  # reorder by count within each facet
  group_by(variable, actor_gender) %>%
  mutate(value = fct_infreq(value)) %>%  # orders factor by descending frequency
  ungroup() %>%
  ggplot(aes(x = value,fill=value)) +
  geom_bar() +
  facet_grid(variable ~ actor_gender, scales = "free_y") +
  labs(title = "Distribution of Actions and Shoulds",
       x = "",
       y = "Count") +
  theme_classic()+
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())+
  scale_fill_brewer(palette = "Pastel1")
```


```{r,fig.height=5,fig.width=8}
library(tidyverse)
library(forcats)

# Prepare plot data
plot_data <- wide %>%
  select(
    id,
    actor_gender,
    would_stranger = stran_act,
    should_stranger = stran_act_should,
    would_acquaintance = acq_act,
    should_acquaintance = acq_act_should,
    would_friend = friend_act,
    should_friend = friend_act_should
  ) %>%
  pivot_longer(
    cols = would_stranger:should_friend,
    names_to = "variable",
    values_to = "value"
  )%>% 
  distinct(id, actor_gender, variable, value) %>% 
  mutate(value = factor(value)) %>%
  group_by(variable, actor_gender) %>%
  mutate(value = fct_infreq(value)) %>%
  ungroup()

# Chi-square test results
test_results <- plot_data %>%
  group_by(variable) %>%
  summarise(
    test = list(chisq.test(table(value, actor_gender))),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      res <- .
      paste0("χ²(", res$parameter, ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )


# Add y-position for label (centered above each facet)
label_positions <- plot_data %>%
  group_by(variable) %>%
  summarise(y_pos = max(table(value)) + 1, .groups = "drop") %>%  # or a fixed number
  left_join(test_results %>% select(variable, label), by = "variable") %>%
  mutate(x = 4.5,y=650)  # center x (approximate middle of the categories)

# Plot
ggplot(plot_data, aes(x = value, fill = actor_gender)) +
  geom_bar(position = position_dodge(width = 0.7)) +
  facet_wrap(~variable, scales = "free_y") +
  geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0.5, hjust = 0.5) +
  labs(title = "Distribution of Actions and Shoulds by Gender",
       x = NULL,
       y = "Count") +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1")+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # rotate 45 degrees
    axis.ticks.x = element_blank()
  )+
   coord_cartesian(ylim = c(0, 700))   # consistent y-axis
```

### Wrongness vs. Appropriateness Ratings by Actor Gender

```{r}
# Summarize mean and SE
summary_stats <- wide %>%
  pivot_longer(cols = c(appropriate, wrong),
               names_to = "judgment", values_to = "rating") %>%
  group_by(actor_gender, judgment) %>%
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    se_rating = sd(rating, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Compute t-test results for each judgment
test_results <- wide %>%
  pivot_longer(cols = c(appropriate, wrong),
               names_to = "judgment", values_to = "rating") %>%
  group_by(judgment) %>%
  summarise(
    t_test = list(t.test(rating ~ actor_gender)),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(t_test, ~{
      res <- .
      paste0("t(", round(res$parameter,1), ") = ", round(res$statistic,2),
             ", p = ", signif(res$p.value,3))
    })
  )

# Determine y position for the label (above bars)
label_positions <- summary_stats %>%
  group_by(judgment) %>%
  summarise(y_pos = max(mean_rating + se_rating) + 0.2, .groups = "drop") %>%
  left_join(test_results %>% select(judgment, label), by = "judgment")

# Bar plot with error bars and test results
ggplot(summary_stats, aes(x = judgment, y = mean_rating, fill = actor_gender)) +
  geom_col(position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(ymin = mean_rating - se_rating,
                    ymax = mean_rating + se_rating),
                width = 0.1,
                position = position_dodge(width = 0.9)) +
  geom_text(data = label_positions,
            aes(x = judgment, y = y_pos, label = label),
            inherit.aes = FALSE,
            vjust = 0) +
  labs(
    x = "Judgment",
    y = "Mean Rating",
    title = "Mean Ratings of Wrongness and Appropriateness by Actor Gender"
  ) +
  theme_classic() +
  scale_y_continuous(limits = c(0, 4))+
  scale_fill_brewer(palette = "Pastel1")
```

## Social distance

### emotion

### action
```{r,fig.height=5,fig.width=8}
library(tidyverse)

# 1 Prepare plot data: merge gender
plot_data <- wide %>%
  select(
    id = ResponseId,
    would_stranger = stran_act,
    should_stranger = stran_act_should,
    would_acquaintance = acq_act,
    should_acquaintance = acq_act_should,
    would_friend = friend_act,
    should_friend = friend_act_should
  ) %>%
  pivot_longer(
    cols = -id,
    names_to = "variable",
    values_to = "value"
  ) %>%
  distinct(id, variable, value) %>%
  mutate(
    value = factor(value),
    measure = if_else(str_detect(variable, "would"), "would", "should"),
    tie_type = case_when(
      str_detect(variable, "friend") ~ "friend",
      str_detect(variable, "acquaintance") ~ "acquaintance",
      str_detect(variable, "stranger") ~ "stranger"
    ),
    tie_type = factor(tie_type, levels = c("friend", "acquaintance", "stranger"))
  )

# 2 Count occurrences for plotting
counts <- plot_data %>%
  group_by(measure, tie_type, value) %>%
  summarise(n = n(), .groups = "drop")

# Global chi-square tests (one per measure)
global_chi <- plot_data %>%
  group_by(measure) %>%
  summarise(
    test = list({
      tbl <- table(value, tie_type)
      if (all(rowSums(tbl) > 0) && all(colSums(tbl) > 0)) chisq.test(tbl) else NULL
    }),
    .groups = "drop"
  ) %>%
  mutate(
    label = map_chr(test, ~{
      if (is.null(.x)) return("Insufficient data")
      paste0("χ²(", .x$parameter, ") = ",
             round(.x$statistic, 2),
             ", p < .001")
    })
  )

# Show results
global_chi 

# 3 Pairwise chi-square comparisons among tie types (within each measure)
pairwise_tests0 <- map_dfr(unique(plot_data$measure), function(m) {
  data_m <- filter(plot_data, measure == m)

  # Only use tie types that actually appear in this measure
  tie_types_m <- unique(data_m$tie_type)

  # Generate all pairwise combinations
  pairs <- combn(tie_types_m, 2, simplify = FALSE)

  map_dfr(pairs, function(pair) {
    sub <- filter(data_m, tie_type %in% pair)
    tbl <- table(sub$value, sub$tie_type)

    # remove rows or columns that are entirely zero
    tbl2 <- tbl[rowSums(tbl) > 0, colSums(tbl) > 0, drop = FALSE]

    if (nrow(tbl2) < 2 || ncol(tbl2) < 2) {
      # skip invalid comparisons
      return(tibble(
        measure = m,
        tie1 = pair[1],
        tie2 = pair[2],
        p_value = NA_real_,
        sig = ""
      ))
    }

    # Run test (safe)
    p <- tryCatch({
      suppressWarnings(chisq.test(tbl2))$p.value
    }, error = function(e) {
      tryCatch({
        fisher.test(tbl2, simulate.p.value = TRUE)$p.value
      }, error = function(e2) NA_real_)
    })

    tibble(
      measure = m,
      tie1 = pair[1],
      tie2 = pair[2],
      p_value = p,
      sig = case_when(
        !is.na(p) & p < 0.001 ~ "***",
        !is.na(p) & p < 0.01  ~ "**",
        !is.na(p) & p < 0.05  ~ "*",
        TRUE ~ ""
      )
    )
  })
})

pairwise_tests0

# 4️⃣ Determine x-positions for significance stars
label_positions <- plot_data %>%
  group_by(measure) %>%
  summarise(y_pos = max(table(value)) + 1, .groups = "drop") %>%  # or a fixed number
  left_join(global_chi %>% select(measure, label), by = "measure") %>%
  mutate(x = 4.5,y=650)  # center x (approximate middle of the categories)

# 5️⃣ Plot
ggplot(counts, aes(x = value, y = n, fill = tie_type)) +
  geom_col(position = position_dodge(width = 0.8)) +
   geom_text(data = label_positions,
            aes(x = x, y = y, label = label),
            inherit.aes = FALSE,
            vjust = 0) +
  facet_wrap(~measure) +
  labs(
    title = "Distribution of Actions by Social Distance",
    x = "Social Distance (Tie Type)",
    y = "Count"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, 750))
```

### emotion 
```{r}
library(tidyverse)

# --- Prepare plot data: count of each emotion by tie type ---
plot_data <- df %>%
  filter(emo_type!="emotion_actor") %>% 
  select(
    id,
    actor_gender,  # optional, can ignore
    emo_type,
    emotion
  )  %>%
  mutate(
    tie_type = case_when(
      emo_type == "emotion_friend" ~ "Friend",
      emo_type == "emotion_acquaintance" ~ "Acquaintance",
      emo_type == "emotion_stranger" ~ "Stranger"
    ),
    tie_type = factor(tie_type, levels = c("Friend", "Acquaintance", "Stranger")),
    emotion = factor(emotion)
  )

# --- Global chi-square test: emotion counts by tie type ---
chi_test <- chisq.test(table(plot_data$emotion, plot_data$tie_type))

chi_label <- paste0(
  "χ²(", chi_test$parameter, ") = ", round(chi_test$statistic, 2),
  ", p < .001"
)

# --- Count occurrences for plotting ---
counts <- plot_data %>%
  group_by(tie_type, emotion) %>%
  summarise(n = n(), .groups = "drop")

# --- Plot ---
ggplot(counts, aes(x = emotion, y = n, fill = tie_type)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    title = "Distribution of Emotions by Tie Type",
    subtitle = chi_label,
    x = NULL,
    y = "Count"
  ) +
  theme_classic() +
  scale_fill_brewer(palette = "Pastel1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, max(counts$n) + 50))

```



# item wise analysis

### gender difference (is there a sig difference for female vs. male for each measure (i.e., emotion, action, wrongness, app) - anova)
```{r}

```


### strong vs. weak tie (is there a sig difference for sanction severeness and emotion intensity for strong tie vs. weak tie - anova)

```{r}

```

### if main effect, interactions?
```{r}


```


# analysis (alignment)

## correlation by category - does LLM predicted human emotional and behavioral responses to norm violation?

### by gender
### by tie strength
### by moral foundation

## difference in alignment by categories









