---
title: "humanxllm-analysis"
format:
    html:
      toc-title: Table of contents
      toc: true
      toc-depth: 2
      code-fold: true # allow code folding
      number-sections: true
      highlight-style: github
      toc-location: body
      cap-location: top
      page-layout: full
      embed-resources: true
      self-contained-math: true
      toc-expand: true
      theme: cosmo
execute:
  echo: false        # show R code
  warning: false    # hide warnings
  message: false    # hide messages
  eval: true        # run the code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F,message = F)
options(download.file.method = "libcurl")

if (!require("groundhog")) install.packages("groundhog")
library("groundhog")
pkgs <- c("tidyverse","car","psych","lme4","table1","jtools","sjPlot","sjmisc","ggpubr","rstatix","lmtest","robustbase","see","rempsyc","sandwich","stargazer","boot","patchwork","data.table","irr","RColorBrewer")
groundhog.library(pkgs, "2025-08-11")

options(scipen = 999, digits = 3)
set.seed(111)
```

```{r}
df<-read.csv( "data/df_all_long_updated.csv") %>% 
  mutate(rater=ifelse(label_type=="human","human","llm")) %>% 
  mutate(unique_id=paste0(label_type,"_",human_id)) %>% 
  mutate(ID=as.factor(ID)) %>% 
  mutate(
  across(
    ends_with(c("_reaction", "_reaction_should")),
    ~ case_when(
      .x %in% c("Celebrate / praise", "Praise") ~ "Praise",
      .x %in% c("Do nothing") ~ "Do nothing",
      .x %in% c("Gossip", "Gossip about / discuss with / complain to others informally") ~ "Gossip",
      .x %in% c("Inform authority", "Inform the authorities, such as the police") ~ "Inform authority",
      .x %in% c("Physical confrontation") ~ "Physical confrontation",
      .x %in% c("Stay away", "Stay away or keep a distance from the actor") ~ "Stay away",
      .x %in% c("Support", "Support the actor") ~ "Support",
      .x %in% c("Verbal confrontation", "Verbal reprimand / verbal confrontation") ~ "Verbal confrontation",
      TRUE ~ .x
    )
  )
) 
  
```

# violin plot with individual response jitter for llm and human
```{r,fig.width=12,fig.height=6}
# Get unique emotion types
emotion_types <- unique(df$emotion_type)

# Create a list of plots, one for each emotion_type
plots <- lapply(emotion_types, function(e_type) {
  df %>%
    filter(emotion_type == e_type) %>%
    ggplot(aes(x = label_type, y = strength_value, fill = rater)) +
    geom_violin(alpha = 0.3, width = 1) +
    geom_jitter(width = 0.2, alpha = 0.1, size = 0.05) +
    facet_wrap(~ emotion, nrow = 2) +
    scale_fill_manual(values = c("lightblue", "salmon")) +
    labs(
      x = "",
      y = "Response Strength",
      title = paste("Human vs LLM Responses:", e_type)
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )
})

# Example: display the first plot
plots[[1]]
```


# correlation of rating for each emotion 

```{r}
# using stimuli as grouping variable if distinct actor gender
# using ID as grouping variable if not distinct actor gender
human_df <- df %>%
  filter(startsWith(unique_id, "human")) %>% 
  group_by(ID, emotion,emotion_type) %>%
  #human median
  #summarise(strength_value = median(strength_value, na.rm = TRUE), .groups = "drop") %>%
  # human mean
  summarise(strength_value = mean(strength_value, na.rm = TRUE), .groups = "drop") %>% 
  mutate(unique_id="human")

llm_df <- df %>%
  filter(!startsWith(unique_id, "human")) %>%
  select(ID, emotion,emotion_type,strength_value,unique_id) %>% 
  group_by(ID,emotion,emotion_type,unique_id) %>% 
  summarise(strength_value = mean(strength_value, na.rm = TRUE), .groups = "drop")
  

combined_df <- bind_rows(human_df, llm_df) %>% 
  pivot_wider(
    names_from = unique_id,
    values_from = strength_value
  ) %>% 
  # fill NA with 0 -> need to get those values
  mutate(across(where(is.numeric), ~ifelse(is.na(.), 0, .)))

# sanity check - is there variance in each groups? -> some groups do not have variance across stimuli
ss<-combined_df %>%
  group_by(emotion, emotion_type) %>%
  summarise(across(where(is.numeric), ~sd(., na.rm = TRUE)))

# when LLM show no variance (perhaps this emotion was not selected) correlation is undefined because the cor equation is cov(X,Y)/sd(Y)sd(Y)

cor_list <- combined_df %>%
  group_by(emotion, emotion_type) %>%
  group_modify(~ {
    # Keep only numeric columns for correlation
    num_df <- .x %>%
      select(where(is.numeric)) 
   # %>% select(where(~sd(., na.rm = TRUE) > 0))  # remove constant columns
    
    mat <- as.matrix(num_df)
    
    # Correlation using pairwise complete observations
    cor_mat <- cor(mat, method = "spearman", use = "pairwise.complete.obs")
    
    # Convert to tidy long format
    cor_melt <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE) %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, Correlation = Freq)
    
    cor_melt
  }) %>%
  ungroup()

```

```{r,fig.width=10,fig.height=5}
# Keep only pairs where one ID is "human" and the other is not
human_cor_list <- cor_list %>%
  filter(
    (unique_id_1 == "human" & unique_id_2 != "human") |
    (unique_id_2 == "human" & unique_id_1 != "human")
  ) %>%
  mutate(unique_id_1=as.character(unique_id_1),
         unique_id_2=as.character(unique_id_2)) %>% 
  mutate(LLM = ifelse(unique_id_1 == "human", unique_id_2, unique_id_1)
  )

# Plot: only human vs. each LLM
ggplot(human_cor_list, aes(x = LLM, y = emotion_type, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(is.na(Correlation), "NA", sprintf("%.2f", Correlation))),
            color = "black", size = 3) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0,na.value = "grey80") +
  facet_wrap(~ emotion,ncol=4) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_blank()
  ) +
  labs(
    title = "Correlation Between Human and LLMs Across Stimuli",
    x = "LLM Model",
    fill = "Correlation"
  )

```

# emotion alignment
##  pairwise comparison- emotion  - euclidean distance
```{r}
# --- compute pairwise similarities
df_wide <- df %>%
  #filter(emotion_type == "emotion_friend") %>% 
  #filter(emotion_type == "emotion_acquaintance") %>% 
  #filter(emotion_type == "emotion_stranger") %>% 
  filter(emotion_type == "emotion_actor") %>% 
  select(ID, stimuli,rater, unique_id, emotion, strength_value,emotion_type) %>% 
  group_by(ID,stimuli,rater, unique_id, emotion,emotion_type) %>% 
  summarise(strength_value = mean(strength_value, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(id_cols = c(unique_id,ID,rater,stimuli),
    names_from = emotion, values_from = strength_value) %>% 
  arrange(ID, unique_id) %>% 
  select(-stimuli) %>% 
  mutate(across(Pride:Compassion, as.numeric))
```

```{r}
# for each stimuli across emotions, compute cosine similarity
pairwise_sim <- df_wide %>%
  group_by(ID) %>%
  do({
    mat <- as.matrix(select(., -rater, -ID,-unique_id))
    rownames(mat) <- .$unique_id
    # compute cosine similarity
    #sim <- as.matrix(proxy::simil(mat, method = "cosine"))
    # compute Euclidean distance
    sim <- as.matrix(proxy::dist(mat, method = "Euclidean"))
    as.data.frame(as.table(sim)) %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, similarity = Freq) %>%
      filter(unique_id_1 != unique_id_2)  # remove self-comparison
  }) %>%
  ungroup() %>% 
  rowwise() %>%
  mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>% 
   mutate(
    unique_id_1_std = ifelse(startsWith(as.character(unique_id_1), "human"), "human", as.character(unique_id_1)),
    unique_id_2_std = ifelse(startsWith(as.character(unique_id_2), "human"), "human", as.character(unique_id_2))
  )

pairwise_sim
```

```{r}
# Compute average  similarity of each non-human ID to humans
avg_sim_to_human <- pairwise_sim %>%
  # Keep only pairs where one side is human and the other is non-human
  filter(
    (unique_id_1_std == "human" & unique_id_2_std != "human") |
    (unique_id_2_std == "human" & unique_id_1_std != "human")
  ) %>%
  # Identify which is the non-human ID in the pair
  mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
  # Group by non-human ID and compute mean  similarity measure
  group_by(non_human_id) %>%
  summarise(avg_similarity_to_human = mean(similarity, na.rm = TRUE)) %>%
  ungroup()

avg_sim_to_human
```

```{r}
# Step 1: Non-human → human average per stimuli × non-human
nonhuman_to_human_avg <- pairwise_sim %>%
  filter((unique_id_1_std == "human" & unique_id_2_std != "human") |
         (unique_id_2_std == "human" & unique_id_1_std != "human")) %>%
  mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
  group_by(ID, non_human_id) %>%
  summarise(
    unique_id_1 = first(unique_id_1_std),
    unique_id_2 = first(unique_id_2_std),
    similarity = mean(similarity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(ID, unique_id_1, unique_id_2, similarity)

# Step 2: Human → human average per stimuli
human_to_human_avg <- pairwise_sim %>%
  filter(unique_id_1_std == "human" & unique_id_2_std == "human") %>%
  group_by(ID) %>%
  summarise(
    unique_id_1 = first(unique_id_1_std),
    unique_id_2 = first(unique_id_2_std),
    similarity = mean(similarity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(ID, unique_id_1, unique_id_2, similarity)

# Step 3: Non-human → non-human (keep original)
nonhuman_to_nonhuman <- pairwise_sim %>%
  filter(unique_id_1_std != "human" & unique_id_2_std != "human") %>%
  select(ID, unique_id_1 = unique_id_1_std, unique_id_2 = unique_id_2_std, similarity)

# Step 4: Combine all into a single table
final_pairwise_sim <- bind_rows(nonhuman_to_human_avg, human_to_human_avg, nonhuman_to_nonhuman)

# remove duplicates in case any exist (e.g., A–B vs B–A)
final_pairwise_sim <- final_pairwise_sim %>%
  mutate(pair_id = paste(pmin(unique_id_1, unique_id_2), pmax(unique_id_1, unique_id_2), sep = "_")) %>%
  distinct(ID, pair_id, .keep_all = TRUE) %>%
  select(-pair_id)
```

### average plot
```{r}
#Compute average
avg_pairwise_sim <- final_pairwise_sim %>%
  mutate(
    # Put the smaller ID in unique_id_1, larger in unique_id_2
    u1 = pmin(unique_id_1, unique_id_2),
    u2 = pmax(unique_id_1, unique_id_2)
  ) %>%
  # Compute average cosine similarity across stimuli
  group_by(u1, u2) %>%
  summarise(avg_similarity = mean(similarity, na.rm = TRUE),
            .groups = "drop") %>% 
  # Reorder u2 so that "human" is last (at the bottom)
  mutate(u2 = factor(u2, levels = c(sort(setdiff(unique(u2), "human")), "human")))


# Plot heatmap with human at top
ggplot(avg_pairwise_sim, aes(x = u1, y = u2, fill = avg_similarity)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "Gn", direction = 1) +
  geom_text(aes(label = round(avg_similarity, 2)), color = "black", size = 4) +

  labs(
    title = "Average Euclidean Distance Across Stimuli",
    x = "",
    y = "",
    fill = "Avg Euclidean Distance"
  ) +
  scale_y_discrete(limits = rev(levels(avg_pairwise_sim$u2))) +  # reverse to put top at top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### linear model
```{r}
df_analysis <- final_pairwise_sim %>%
  # ensure character (important if columns are factors)
  mutate(
    u1 = as.character(unique_id_1),
    u2 = as.character(unique_id_2)
  ) %>%
  # create order-invariant pair id (vectorized)
  mutate(
    pair_model = paste(pmin(u1, u2), pmax(u1, u2), sep = "_"),
    pair_type = case_when(
      u1 == "human" & u2 == "human" ~ "human_human",
      u1 == "human" | u2 == "human" ~ "LLM_human",
      TRUE ~ "LLM_LLM"
    )
  ) %>%
  # optional: keep only human-containing pairs
  filter(pair_type != "LLM_LLM") %>%
  select(-u1, -u2)  %>% 
  mutate(
    # Make "human_human" the reference level for both
    pair_type = factor(pair_type, levels = c("human_human", "LLM_human")),
    pair_model = factor(pair_model, levels = c("human_human", setdiff(unique(pair_model), "human_human")))
  )

# overall
model<-lmer(similarity~pair_type + (1|ID), df_analysis)
# each model
model2<-lmer(similarity~pair_model + (1|ID), df_analysis)
tab_model(model,model2)
```

## similarity for each emotions

```{r}
# Compute pairwise differences using only the 'Disgust' column
pairwise_dist <- df_wide %>%
  group_by(ID) %>%
  do({
    ids <- .$unique_id
    values <- .$Disgust  # numeric column

    # Compute pairwise differences (not absolute)
    dist_mat <- outer(values, values, FUN = "-")

    # Convert to long format
    dist_df <- as.data.frame(as.table(dist_mat))
    dist_df <- dist_df %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, distance = Freq) %>%
      filter(unique_id_1 != unique_id_2)

    # Restore proper unique_id labels
    dist_df$unique_id_1 <- ids[as.numeric(dist_df$unique_id_1)]
    dist_df$unique_id_2 <- ids[as.numeric(dist_df$unique_id_2)]

    dist_df
  }) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>%
  mutate(
    unique_id_1_std = ifelse(startsWith(unique_id_1, "human"), "human", unique_id_1),
    unique_id_2_std = ifelse(startsWith(unique_id_2, "human"), "human", unique_id_2)
  )

pairwise_dist
```

```{r}
# List of emotion columns
emotions <- c("Pride","Anger","Guilt","Shame","Embarrassment","Disgust","Contempt","Compassion")

# Initialize empty tables to store results
all_pairwise <- list()
all_models <- list()
all_coefs <- tibble()
all_avg <- list()

for(em in emotions){
  
  # Step 1: Pairwise differences=============================================
  pairwise_dist <- df_wide %>%
    group_by(ID) %>%
    do({
      ids <- .$unique_id
      values <- .[[em]]  
      # Compute absolute pairwise differences
      dist_mat <- outer(values, values, FUN = function(x, y) abs(x - y))
    
      dist_df <- as.data.frame(as.table(dist_mat)) %>%
        rename(unique_id_1 = Var1, unique_id_2 = Var2, distance = Freq) %>%
        filter(unique_id_1 != unique_id_2)
      
      dist_df$unique_id_1 <- ids[as.numeric(dist_df$unique_id_1)]
      dist_df$unique_id_2 <- ids[as.numeric(dist_df$unique_id_2)]
      dist_df
    }) %>%
    ungroup() %>%
    rowwise() %>%
    mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
    distinct(pair, .keep_all = TRUE) %>%
    select(-pair) %>%
    mutate(
      unique_id_1_std = ifelse(startsWith(unique_id_1, "human"), "human", unique_id_1),
      unique_id_2_std = ifelse(startsWith(unique_id_2, "human"), "human", unique_id_2),
      emotion = em
    )
  
  # Step 2: Aggregate by type=============================================
  nonhuman_to_human_avg <- pairwise_dist %>%
    filter((unique_id_1_std == "human" & unique_id_2_std != "human") |
           (unique_id_2_std == "human" & unique_id_1_std != "human")) %>%
    mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
    group_by(ID, non_human_id) %>%
    summarise(
      unique_id_1 = first(unique_id_1_std),
      unique_id_2 = first(unique_id_2_std),
      distance = mean(distance, na.rm = TRUE),
      emotion = first(emotion),
      .groups = "drop"
    ) %>%
    select(ID, unique_id_1, unique_id_2, distance, emotion)
  
  human_to_human_avg <- pairwise_dist %>%
    filter(unique_id_1_std == "human" & unique_id_2_std == "human") %>%
    group_by(ID) %>%
    summarise(
      unique_id_1 = first(unique_id_1_std),
      unique_id_2 = first(unique_id_2_std),
      distance = mean(distance, na.rm = TRUE),
      emotion = first(emotion),
      .groups = "drop"
    )
  
  nonhuman_to_nonhuman <- pairwise_dist %>%
    filter(unique_id_1_std != "human" & unique_id_2_std != "human") %>%
    select(ID, unique_id_1 = unique_id_1_std, unique_id_2 = unique_id_2_std, distance) %>%
    mutate(emotion = em)
  
  final_pairwise_dist <- bind_rows(nonhuman_to_human_avg, human_to_human_avg, nonhuman_to_nonhuman) %>%
    mutate(pair_id = paste(pmin(unique_id_1, unique_id_2), pmax(unique_id_1, unique_id_2), sep = "_")) %>%
    distinct(ID, pair_id, .keep_all = TRUE) %>%
    select(-pair_id)
  
  # Store results
  all_pairwise[[em]] <- final_pairwise_dist
  
  # Step 3: Linear model for human vs non-human=============================================
  df_analysis <- final_pairwise_dist %>%
    mutate(u1 = as.character(unique_id_1),
           u2 = as.character(unique_id_2),
           pair_model = paste(pmin(u1, u2), pmax(u1, u2), sep = "_"),
           pair_type = case_when(
             u1 == "human" & u2 == "human" ~ "human_human",
             u1 == "human" | u2 == "human" ~ "LLM_human",
             TRUE ~ "LLM_LLM"
           )) %>%
    filter(pair_type != "LLM_LLM") %>%
    mutate(pair_type = factor(pair_type, levels = c("human_human", "LLM_human")),
           pair_model = factor(pair_model))
  
  model <- lmer(distance ~ pair_type + (1|ID), df_analysis)
  model2 <- lmer(distance ~ pair_model + (1|ID), df_analysis)
  
  all_models[[em]] <- list(model=model, model2=model2)
  # Extract coefficients with broom.mixed::tidy()
    coefs1 <- tidy(model) %>% mutate(emotion = em, model_type = "pair_type")
    coefs2 <- tidy(model2) %>% mutate(emotion = em, model_type = "pair_model")
    all_coefs <- bind_rows(all_coefs, coefs1, coefs2)
    
  # Step 4: Average across stimuli=============================================
  avg_pairwise_dist <- final_pairwise_dist %>%
    mutate(u1 = pmin(unique_id_1, unique_id_2),
           u2 = pmax(unique_id_1, unique_id_2)) %>%
    group_by(u1, u2, emotion) %>%
    summarise(avg_distance = mean(distance, na.rm = TRUE), .groups = "drop") %>%
    mutate(u2 = factor(u2, levels = c(sort(setdiff(unique(u2), "human")), "human")))
  
  # Store results
  all_avg[[em]] <- avg_pairwise_dist
}

# Select relevant columns for a clean table =============================================
all_coefs %>%
  filter(!is.na(statistic), term != "(Intercept)") %>%
  mutate(
    p.value = 2 * (1 - pnorm(abs(statistic))),
    stars = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ),
    # Compute 95% CI
    ci_lower = estimate - 1.96 * std.error,
    ci_upper = estimate + 1.96 * std.error,
    # Combine into single string: estimate*** (CI)
    est_ci = sprintf("%.2f%s [%.2f, %.2f]", estimate, stars, ci_lower, ci_upper)
  ) %>%
  select(emotion, model_type, term, est_ci)%>% 
  pivot_wider(
    names_from = emotion,
    values_from = est_ci
  ) %>% 
  nice_table(.)
```

```{r,fig.height=6,fig.width=8}
# Combine all emotions into single long tables
combined_pairwise <- bind_rows(all_pairwise)
combined_avg <- bind_rows(all_avg)
  

ggplot(combined_avg, aes(x = u1, y = u2, fill = avg_distance)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(avg_distance, 2)), size = 3) +
  scale_fill_distiller(palette = "Gn", direction = 1) +
  scale_y_discrete(limits = levels(combined_avg$u2)) +  # DO NOT reverse, keep human first
  labs(
    title = "Average Emotion Differences Across Stimuli",
    x = "", y = "", fill = "Avg Distance"
  ) +
  facet_wrap(~emotion) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r,eval=F}
# correlation across stimuli 
combined_pairwise <- bind_rows(all_pairwise)

# Pivot to wide format: each row = stimulus, columns = pairs
pair_matrix <- combined_pairwise %>%
  mutate(pair = paste(unique_id_1, unique_id_2, sep = "_")) %>%
  select(ID, pair, distance,emotion) %>%
  group_by(ID, pair,emotion) %>% 
  summarise(distance = mean(distance),.groups = "drop") %>% 
  pivot_wider(names_from = pair, values_from = distance, values_fill = list(distance = NA))

# Convert to numeric matrix
stim_matrix <- as.matrix(pair_matrix[,-1])
rownames(stim_matrix) <- pair_matrix$ID
# Correlation between rows (stimuli) based on all pairs
cor_matrix <- cor(t(stim_matrix), use = "pairwise.complete.obs")

cor_melt <- melt(cor_matrix)
colnames(cor_melt) <- c("Stimulus1", "Stimulus2", "Correlation")

ggplot(cor_melt, aes(x = Stimulus1, y = Stimulus2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Across Stimuli (All Pairs)",
       x = "Stimulus",
       y = "Stimulus")
```


# action alignment

## pariwise comparison
```{r}
# -----------------------------
# 1. Select relevant action variables
# -----------------------------
action_vars <- c(
  "strangers_reaction",
  "acquaintances_reaction",
  "friends_reaction",
  "strangers_reaction_should",
  "acquaintances_reaction_should",
  "friends_reaction_should"
)

df_action <- df %>%
  select(unique_id, ID, all_of(action_vars)) %>%
  distinct()

# -----------------------------
# 2. Reshape into long format
# -----------------------------
df_action_long <- df_action %>%
  pivot_longer(
    cols = -c(unique_id, ID),
    names_to = "action_var",
    values_to = "choice"
  ) %>%
  mutate(
    # identify social distance
    social_distance = case_when(
      grepl("strangers", action_var) ~ "stranger",
      grepl("acquaintances", action_var) ~ "acquaintance",
      grepl("friends", action_var) ~ "friend"
    ),
    # identify action type
    action_type = ifelse(
      grepl("_should$", action_var),
      "should",
      "would"
    )
  ) %>%
  select(unique_id, ID, social_distance, action_type, choice) %>% 
  # drop 18 missing 
  filter(!is.na(choice))

# -----------------------------
# 3. Compute pairwise choice agreement (0/1)
# -----------------------------
pairwise_action_sim <- df_action_long %>%
  group_by(ID, social_distance, action_type) %>%
  group_modify(~{
    dat <- .x
    # all unique pairs of agents
    pairs <- combn(seq_len(nrow(dat)), 2, simplify = FALSE)
    map_dfr(pairs, function(idx) {
      a <- dat[idx[1], ]
      b <- dat[idx[2], ]
      tibble(
        unique_id_1 = a$unique_id,
        unique_id_2 = b$unique_id,
        agreement = as.integer(a$choice == b$choice)
      )
    })
  }) %>%
  ungroup() %>%
  mutate(
    # standardize human IDs
    unique_id_1_std = ifelse(startsWith(unique_id_1, "human"), "human", unique_id_1),
    unique_id_2_std = ifelse(startsWith(unique_id_2, "human"), "human", unique_id_2),
    # classify pair type
    pair_type = case_when(
      unique_id_1_std == "human" & unique_id_2_std == "human" ~ "human_human",
      unique_id_1_std == "human" | unique_id_2_std == "human" ~ "LLM_human",
      TRUE ~ "LLM_LLM"
    ),
    # order-invariant pair ID
    pair_id = paste(pmin(unique_id_1, unique_id_2), pmax(unique_id_1, unique_id_2), sep = "_")
  ) %>%
  distinct(ID, social_distance, action_type, pair_id, .keep_all = TRUE)

# -----------------------------
# 4. Function to fit logistic mixed-effects model per condition
# -----------------------------
fit_action_model <- function(data, distance, action) {
  dat <- data %>%
    filter(
      social_distance == distance,
      action_type == action,
      pair_type != "LLM_LLM" # exclude LLM–LLM
    ) %>%
    mutate(pair_type = factor(pair_type, levels = c("human_human", "LLM_human")))
  
  glmer(
    agreement ~ pair_type + (1 | ID),
    data = dat,
    family = binomial
  )
}

# -----------------------------
# 5. Fit models for each social distance × action type
# -----------------------------
models_action <- list(
  stranger_would = fit_action_model(pairwise_action_sim, "stranger", "would"),
  stranger_should = fit_action_model(pairwise_action_sim, "stranger", "should"),
  acquaintance_would = fit_action_model(pairwise_action_sim, "acquaintance", "would"),
  acquaintance_should = fit_action_model(pairwise_action_sim, "acquaintance", "should"),
  friend_would = fit_action_model(pairwise_action_sim, "friend", "would"),
  friend_should = fit_action_model(pairwise_action_sim, "friend", "should")
)

# -----------------------------
# 6. Extract results in tidy format
# -----------------------------
action_results <- purrr::imap_dfr(
  models_action,
  ~ tidy(.x, effects = "fixed") %>%
      filter(term == "pair_typeLLM_human") %>%
      mutate(
        condition = .y,
        odds_ratio = exp(estimate)
      )
)

action_results

# -----------------------------
# 7. Optional: Descriptive agreement rates
# -----------------------------
agreement_rates <- pairwise_action_sim %>%
  filter(pair_type != "LLM_LLM") %>%
  group_by(social_distance, action_type, pair_type) %>%
  summarise(
    agreement_rate = mean(agreement),
    .groups = "drop"
  )

agreement_rates
```
```{r}
# Reorder factors for nicer plotting
df_agreement <- agreement_rates %>%
  mutate(
    social_distance = factor(social_distance, levels = c("stranger", "acquaintance", "friend")),
    action_type = factor(action_type, levels = c("would", "should")),
    pair_type = factor(pair_type, levels = c("human_human", "LLM_human"))
  )

library(RColorBrewer)

ggplot(df_agreement, aes(x = pair_type, y = social_distance, fill = agreement_rate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = scales::percent(agreement_rate, accuracy = 1)), color = "black") +
  facet_wrap(~action_type) +
  scale_fill_distiller(palette = "Blues", direction = 1, name = "Agreement Rate") +
  labs(
    title = "",
    x = "Pair Type",
    y = "Social Distance"
  ) +
  theme_minimal(base_size = 14)
```

# Do emotion predict enforcement?

```{r}
actions <- c(
  "do_nothing",
  "physical_confrontation",
  "verbal_confrontation",
  "inform_authority",
  "gossip",
  "stay_away",
  "praise",
  "support"
)

# flat all action to binary
df_action <- df %>%
  mutate(
    across(
      c(friends_reaction, strangers_reaction, acquaintances_reaction),
      .fns = setNames(
        lapply(actions, function(a) {
          function(x) as.integer(x == a)
        }),
        paste0("is_", gsub(" ", "_", tolower(actions)))
      ),
      .names = "{.col}_{.fn}"
    )
  ) %>% 
  filter(emotion_type!="emotion_actor") %>% 
  rename_with(
    ~ gsub(" ", "_", tolower(.x)),
    contains("_reaction_is_")
  )

action_map <- tibble::tibble(
  action_label = c(
    "Do nothing",
    "Physical confrontation",
    "Verbal confrontation",
    "Inform authority",
    "Gossip",
    "Stay away",
    "Praise",
    "Support"
  ),
  action = gsub(" ", "_", tolower(action_label))
)

combinations <- tidyr::expand_grid(
  emotion = unique(df_action$emotion),
  emotion_type = unique(df_action$emotion_type),
  action = action_map$action
) %>% 
  filter(emotion_type!="emotion_actor")
```

```{r}
fit_model_safe <- function(df, outcome_col) {
  # Check if outcome has both 0 and 1
  if(length(unique(na.omit(df[[outcome_col]]))) < 2) return(NULL)
  glm(as.formula(paste0(outcome_col, " ~ strength_value * rater")), 
      data = df, family = binomial)
}

models <- pmap(
  combinations,
  function(emotion, emotion_type, action) {

    outcome_col <- case_when(
      emotion_type == "emotion_friend" ~
        paste0("friends_reaction_is_", action),
      emotion_type == "emotion_acquaintance" ~
        paste0("acquaintances_reaction_is_", action),
      emotion_type == "emotion_stranger" ~
        paste0("strangers_reaction_is_", action)
    )

    df_sub <- df_action %>%
      filter(emotion == !!emotion,
             emotion_type == !!emotion_type)

    model <- fit_model_safe(df_sub, outcome_col)

    list(
      emotion = emotion,
      emotion_type = emotion_type,
      action = action,
      outcome = outcome_col,
      model = model
    )
  }
)

# Remove NULL models
models <- keep(models, ~ !is.null(.x$model))
```

```{r}
library(broom)
library(purrr)
library(dplyr)

models_clean <- map(models, "model")

model_results <- map_dfr(models, function(x) {

  tidy_df <- tidy(x$model)
  if (nrow(tidy_df) == 0) return(NULL)

  tidy_df %>%
    mutate(
      emotion = x$emotion,
      emotion_type = x$emotion_type,
      action = x$action,
      odds_ratio = exp(estimate)
    )
})

```

```{r}
# Extract glm objects if nested
models_clean <- map(models, ~ if(is.list(.x) && "model" %in% names(.x)) .x$model else .x)

# Remove NULLs
models_clean <- keep(models_clean, ~ !is.null(.x))

# Extract tidy results
model_results <- imap_dfr(models_clean, ~ {
  tidy(.x) %>%
    mutate(
      model_name = .y,
      odds_ratio = exp(estimate)
    )
})

# Round numbers for readability
model_results <- model_results %>%
  select(model_name, term, estimate, std.error, statistic, p.value, odds_ratio) %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = round(statistic, 2),
    p.value = signif(p.value, 3),
    odds_ratio = round(odds_ratio, 2)
  )

# Generate LaTeX table using knitr
#knitr::kable(
#  model_results,
#  format = "latex",
#  booktabs = TRUE,
#  caption = "GLM model results for approach/avoidance reactions by emotion and social distance",
#  col.names = c("Model", "Term", "Estimate", "SE", "Z", "p", "Odds Ratio")
#)
```

```{r}
pred_list <- lapply(models, function(x) {

  model <- x$model
  if (is.null(model)) return(NULL)

  newdata <- expand.grid(
    strength_value = 0:4,
    rater = unique(model$data$rater)
  )

  pred_link <- predict(model, newdata, type = "link", se.fit = TRUE)

  newdata$predicted <- plogis(pred_link$fit)
  newdata$conf.low  <- plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata$conf.high <- plogis(pred_link$fit + 1.96 * pred_link$se.fit)

  newdata$emotion <- x$emotion
  newdata$emotion_type <- x$emotion_type
  newdata$action <- x$action          
  newdata$outcome <- x$outcome       

  newdata
})

pred_combined <- bind_rows(pred_list)

```

```{r,fig.height=10,fig.width=10}
# Optional: prettier labels for facets
pred_combined <- pred_combined %>%
  mutate(
    action_label = gsub("_", " ", action),     
    #emotion_short = abbreviate(emotion, minlength = 4)
  )

table(pred_combined$emotion_type)

ggplot(
  filter(pred_combined,emotion_type =="emotion_acquaintance"),
  aes(
    x = strength_value,
    y = predicted,
    color = rater,
    fill = rater
  )
) +
  geom_line(size = 1.2) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.25,
    color = NA
  ) +
  facet_grid(
    action ~ emotion,
    scales = "free_y",
    switch = "y"
  ) +
  labs(
    x = "Emotion Strength",
    y = "Predicted Probability of Action",
    color = "Rater",
    fill = "Rater",
    title = "emotion_acquaintance"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    axis.text.x = element_text(size = 11, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    panel.spacing = unit(0.25, "lines"),
    strip.background = element_rect(fill = "gray95")
  ) +
  guides(
    color = guide_legend(nrow = 1),
    fill = guide_legend(nrow = 1)
  )

```


## old approach avoidance
```{r, eval=F}
approach <- c("Inform authority","Physical confrontation", "Verbal confrontation" )
avoidance <- c("Gossip", "Stay away","Do nothing",  "Praise", "Support")


df_action <- df %>%
  mutate(across(c(friends_reaction, strangers_reaction, acquaintances_reaction),
                ~ case_when(
                  .x %in% approach ~ "approach",
                  .x %in% avoidance ~ "avoidance",
                  TRUE ~ NA_character_
                ),
                .names = "{.col}_category")) %>% 
  mutate(across(c(friends_reaction, strangers_reaction, acquaintances_reaction),
                ~ ifelse(.x %in% approach, 1, 0),
                .names = "{.col}_binary"))

# -----------------------------
# 2. Function to fit model safely
# -----------------------------
fit_model_safe <- function(df, outcome_col) {
  # Check if outcome has both 0 and 1
  if(length(unique(na.omit(df[[outcome_col]]))) < 2) return(NULL)
  glm(as.formula(paste0(outcome_col, " ~ strength_value * rater")), 
      data = df, family = binomial)
}

# -----------------------------
# 3. Fit models for all emotion x emotion_type x social distance
# -----------------------------
# Define combinations
combinations <- expand.grid(
  emotion = unique(df_action$emotion),
  emotion_type = unique(df_action$emotion_type),
  stringsAsFactors = FALSE
)

# Fit models dynamically
models <- pmap(
  combinations,
  function(emotion, emotion_type) {
    # Determine which column to model based on emotion_type
    outcome_col <- case_when(
      emotion_type == "emotion_friend" ~ "friends_reaction_binary",
      emotion_type == "emotion_acquaintance" ~ "acquaintances_reaction_binary",
      emotion_type == "emotion_stranger" ~ "strangers_reaction_binary"
    )
    
    df_sub <- df_action %>%
      filter(emotion == !!emotion, emotion_type == !!emotion_type)
    
    model <- fit_model_safe(df_sub, outcome_col)
    
    list(emotion = emotion,
         emotion_type = emotion_type,
         model = model)
  }
)

# Remove NULLs (cases with only 0 or 1)
models <- keep(models, ~ !is.null(.x$model))

library(broom)
library(purrr)
```

```{r,eval=F}
# -----------------------------
# 4. Generate predicted probabilities
# -----------------------------
pred_list <- lapply(models, function(x) {
  model <- x$model
  if(is.null(model)) return(NULL)
  
  newdata <- expand.grid(
    strength_value = 0:4,
    rater = unique(model$data$rater)
  )
  
  newdata$predicted <- predict(model, newdata, type = "response")
  pred_link <- predict(model, newdata, type = "link", se.fit = TRUE)
  newdata$conf.low <- plogis(pred_link$fit - 1.96 * pred_link$se.fit)
  newdata$conf.high <- plogis(pred_link$fit + 1.96 * pred_link$se.fit)
  newdata$emotion <- x$emotion
  newdata$emotion_type <- x$emotion_type
  return(newdata)
})

pred_combined <- bind_rows(pred_list)

# -----------------------------
# 5. Plot predicted probabilities dynamically
# -----------------------------
library(grid)
# Optional: abbreviate emotion names to save space
pred_combined <- pred_combined %>%
  mutate(emotion_type= abbreviate(sub(".*_", "", emotion_type),minlength = 4),
    emotion_short = abbreviate(emotion, minlength = 4))

ggplot(pred_combined, aes(x = strength_value, y = predicted, color = rater, fill = rater)) +
  geom_line(size = 1.5) +                       # thicker lines
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.25, color = NA) +
  facet_grid(emotion_type ~ emotion_short, scales = "free_y", switch = "y") + # emotions in columns, types in rows
  scale_fill_manual(values = c("orange", "skyblue")) +
  scale_color_manual(values = c("orange", "skyblue")) +
  labs(
    x = "Emotion Strength",
    y = "Predicted Probability of Approach",
    color = "Rater",
    fill = "Rater"
  ) +
  theme_classic(base_size = 14) +               # much bigger text
  theme(
    strip.text = element_text(size = 12, face = "bold"),  # facet labels
    axis.text.x = element_text(size = 12, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    legend.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm"),
    legend.key.height = unit(0.8, "cm"),
    panel.spacing = unit(0.2, "lines"),        # less empty space between facets
    strip.background = element_rect(fill = "gray95"),
    plot.margin = margin(5, 5, 5, 5)           # minimize outer margins
  ) +
  guides(
    color = guide_legend(nrow = 1, byrow = TRUE),
    fill = guide_legend(nrow = 1, byrow = TRUE)
  )

```

# alignment analysis 

## correlation by category - does LLM predicted human emotional and behavioral responses to norm violation?

### by gender
### by tie strength
### by moral foundation

## difference in alignment by categories

# item wise analysis

### gender difference (is there a sig difference for female vs. male for each measure (i.e., emotion, action, wrongness, app) - anova)

### strong vs. weak tie (is there a sig difference for sanction severeness and emotion intensity for strong tie vs. weak tie - anova)

### if main effect, interactions?




