---
title: "humanxllm-analysis"
format:
    html:
      toc-title: Table of contents
      toc: true
      toc-depth: 2
      code-fold: true # allow code folding
      number-sections: true
      highlight-style: github
      toc-location: body
      cap-location: top
      page-layout: full
      embed-resources: true
      self-contained-math: true
      toc-expand: true
      theme: cosmo
execute:
  echo: false        # show R code
  warning: false    # hide warnings
  message: false    # hide messages
  eval: true        # run the code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning=F,message = F)
options(download.file.method = "libcurl")

if (!require("groundhog")) install.packages("groundhog")
library("groundhog")
pkgs <- c("tidyverse","car","psych","lme4","table1","jtools","sjPlot","sjmisc","ggpubr","rstatix","lmtest","robustbase","see","rempsyc","sandwich","stargazer","boot","patchwork","data.table","irr","RColorBrewer")
groundhog.library(pkgs, "2025-08-11")

options(scipen = 999, digits = 3)
set.seed(111)
```

```{r}
df<-read.csv( "data/df_all_long_updated.csv") %>% 
  mutate(rater=ifelse(label_type=="human","human","llm")) %>% 
  mutate(unique_id=paste0(label_type,"_",human_id))
```

```{r}
# --- compute pairwise similarities
df_wide <- df %>%
  #filter(emotion_type == "emotion_friend") %>% 
  #filter(emotion_type == "emotion_acquaintance") %>% 
  #filter(emotion_type == "emotion_stranger") %>% 
  filter(emotion_type == "emotion_actor") %>% 
  select(id=ID, rater, unique_id,stimuli, emotion, strength_value,emotion_type) %>% 
  pivot_wider(id_cols = c(unique_id,stimuli,rater),
    names_from = emotion, values_from = strength_value) %>% 
  arrange(stimuli, unique_id) %>% 
  mutate(across(Pride:Compassion, as.numeric))
```

# correlation of rating for each emotion 
```{r}
human_df <- df %>%
  filter(startsWith(unique_id, "human")) %>% 
  group_by(stimuli, emotion,emotion_type) %>%
  summarise(strength_value = median(strength_value, na.rm = TRUE), .groups = "drop") %>% 
  mutate(unique_id="human")

llm_df <- df %>%
  filter(!startsWith(unique_id, "human")) %>%
  select(stimuli, emotion,emotion_type,strength_value,unique_id)

combined_df <- bind_rows(human_df, llm_df) %>% 
  pivot_wider(
    names_from = unique_id,
    values_from = strength_value
  ) %>%
  ungroup()

cor_list <- combined_df %>%
  group_by(emotion, emotion_type) %>%
  group_modify(~ {
    # Keep only numeric columns for correlation
    num_df <- .x %>%
      select(where(is.numeric))
    
    mat <- as.matrix(num_df)
    
    # Correlation using pairwise complete observations
    cor_mat <- cor(mat, use = "pairwise.complete.obs")
    
    # Convert to tidy long format
    cor_melt <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE) %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, Correlation = Freq)
    
    cor_melt
  }) %>%
  ungroup()



```

```{r,fig.width=10,fig.height=5}
# Keep only pairs where one ID is "human" and the other is not
human_cor_list <- cor_list %>%
  filter(
    (unique_id_1 == "human" & unique_id_2 != "human") |
    (unique_id_2 == "human" & unique_id_1 != "human")
  ) %>%
  mutate(unique_id_1=as.character(unique_id_1),
         unique_id_2=as.character(unique_id_2)) %>% 
  mutate(LLM = ifelse(unique_id_1 == "human", unique_id_2, unique_id_1)
  )

# Plot: only human vs. each LLM
ggplot(human_cor_list, aes(x = LLM, y = emotion_type, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), size = 3) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  facet_wrap(~ emotion,ncol=4) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y = element_blank()
  ) +
  labs(
    title = "Correlation Between Human and LLMs Across Stimuli",
    x = "LLM Model",
    fill = "Correlation"
  )

```


# cosine similarity measure 
```{r}
# for each stimuli across emotions, compute cosine similarity
pairwise_cos <- df_wide %>%
  group_by(stimuli) %>%
  do({
    mat <- as.matrix(select(., -unique_id, -stimuli))
    rownames(mat) <- .$unique_id
    cos_sim <- as.matrix(proxy::simil(mat, method = "cosine"))
    as.data.frame(as.table(cos_sim)) %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, cosine_similarity = Freq) %>%
      filter(unique_id_1 != unique_id_2)  # remove self-comparison
  }) %>%
  ungroup() %>% 
  rowwise() %>%
  mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>% 
   mutate(
    unique_id_1_std = ifelse(startsWith(as.character(unique_id_1), "human"), "human", as.character(unique_id_1)),
    unique_id_2_std = ifelse(startsWith(as.character(unique_id_2), "human"), "human", as.character(unique_id_2))
  )

pairwise_cos
```

```{r}
# Compute average cosine similarity of each non-human ID to humans
avg_cosine_to_human <- pairwise_cos %>%
  # Keep only pairs where one side is human and the other is non-human
  filter(
    (unique_id_1_std == "human" & unique_id_2_std != "human") |
    (unique_id_2_std == "human" & unique_id_1_std != "human")
  ) %>%
  # Identify which is the non-human ID in the pair
  mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
  # Group by non-human ID and compute mean cosine similarity
  group_by(non_human_id) %>%
  summarise(avg_cosine_similarity_to_human = mean(cosine_similarity, na.rm = TRUE)) %>%
  ungroup()

avg_cosine_to_human
```

```{r}
# Step 1: Non-human → human average per stimuli × non-human
nonhuman_to_human_avg <- pairwise_cos %>%
  filter((unique_id_1_std == "human" & unique_id_2_std != "human") |
         (unique_id_2_std == "human" & unique_id_1_std != "human")) %>%
  mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
  group_by(stimuli, non_human_id) %>%
  summarise(
    unique_id_1 = first(unique_id_1_std),
    unique_id_2 = first(unique_id_2_std),
    cosine_similarity = mean(cosine_similarity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(stimuli, unique_id_1, unique_id_2, cosine_similarity)

# Step 2: Human → human average per stimuli
human_to_human_avg <- pairwise_cos %>%
  filter(unique_id_1_std == "human" & unique_id_2_std == "human") %>%
  group_by(stimuli) %>%
  summarise(
    unique_id_1 = first(unique_id_1_std),
    unique_id_2 = first(unique_id_2_std),
    cosine_similarity = mean(cosine_similarity, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  select(stimuli, unique_id_1, unique_id_2, cosine_similarity)

# Step 3: Non-human → non-human (keep original)
nonhuman_to_nonhuman <- pairwise_cos %>%
  filter(unique_id_1_std != "human" & unique_id_2_std != "human") %>%
  select(stimuli, unique_id_1 = unique_id_1_std, unique_id_2 = unique_id_2_std, cosine_similarity)

# Step 4: Combine all into a single table
final_pairwise_cos <- bind_rows(nonhuman_to_human_avg, human_to_human_avg, nonhuman_to_nonhuman)

# Optional: remove duplicates in case any exist (e.g., A–B vs B–A)
final_pairwise_cos <- final_pairwise_cos %>%
  mutate(pair_id = paste(pmin(unique_id_1, unique_id_2), pmax(unique_id_1, unique_id_2), sep = "_")) %>%
  distinct(stimuli, pair_id, .keep_all = TRUE) %>%
  select(-pair_id)
```

### average plot
```{r}
#Compute average
avg_pairwise_cos <- final_pairwise_cos %>%
  mutate(
    # Put the smaller ID in unique_id_1, larger in unique_id_2
    u1 = pmin(unique_id_1, unique_id_2),
    u2 = pmax(unique_id_1, unique_id_2)
  ) %>%
  # Compute average cosine similarity across stimuli
  group_by(u1, u2) %>%
  summarise(avg_cosine_similarity = mean(cosine_similarity, na.rm = TRUE),
            .groups = "drop") %>% 
  # Reorder u2 so that "human" is last (at the bottom)
  mutate(u2 = factor(u2, levels = c(sort(setdiff(unique(u2), "human")), "human")))

avg_pairwise_cos

# Plot heatmap with human at top
ggplot(avg_pairwise_cos, aes(x = u1, y = u2, fill = avg_cosine_similarity)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "Gn", direction = 1) +
  geom_text(aes(label = round(avg_cosine_similarity, 2)), color = "black", size = 4) +

  labs(
    title = "Average Cosine Similarity Across Stimuli",
    x = "Unique ID 1",
    y = "Unique ID 2",
    fill = "Avg Cosine"
  ) +
  scale_y_discrete(limits = rev(levels(avg_pairwise_cos$u2))) +  # reverse to put top at top
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

### linear model
```{r}
df_analysis <- final_pairwise_cos %>%
  # ensure character (important if columns are factors)
  mutate(
    u1 = as.character(unique_id_1),
    u2 = as.character(unique_id_2)
  ) %>%
  # create order-invariant pair id (vectorized)
  mutate(
    pair_model = paste(pmin(u1, u2), pmax(u1, u2), sep = "_"),
    pair_type = case_when(
      u1 == "human" & u2 == "human" ~ "human_human",
      u1 == "human" | u2 == "human" ~ "LLM_human",
      TRUE ~ "LLM_LLM"
    )
  ) %>%
  # optional: keep only human-containing pairs
  filter(pair_type != "LLM_LLM") %>%
  select(-u1, -u2)  %>% 
  mutate(
    # Make "human_human" the reference level for both
    pair_type = factor(pair_type, levels = c("human_human", "LLM_human")),
    pair_model = factor(pair_model, levels = c("human_human", setdiff(unique(pair_model), "human_human")))
  )

# overall
model<-lmer(cosine_similarity~pair_type + (1|stimuli), df_analysis)
# each model
model2<-lmer(cosine_similarity~pair_model + (1|stimuli), df_analysis)
tab_model(model,model2)
```

# similarity for each emotions

```{r}
# Compute pairwise differences using only the 'Disgust' column
pairwise_dist <- df_wide %>%
  group_by(stimuli) %>%
  do({
    ids <- .$unique_id
    values <- .$Disgust  # numeric column

    # Compute pairwise differences (not absolute)
    dist_mat <- outer(values, values, FUN = "-")

    # Convert to long format
    dist_df <- as.data.frame(as.table(dist_mat))
    dist_df <- dist_df %>%
      rename(unique_id_1 = Var1, unique_id_2 = Var2, distance = Freq) %>%
      filter(unique_id_1 != unique_id_2)

    # Restore proper unique_id labels
    dist_df$unique_id_1 <- ids[as.numeric(dist_df$unique_id_1)]
    dist_df$unique_id_2 <- ids[as.numeric(dist_df$unique_id_2)]

    dist_df
  }) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) %>%
  mutate(
    unique_id_1_std = ifelse(startsWith(unique_id_1, "human"), "human", unique_id_1),
    unique_id_2_std = ifelse(startsWith(unique_id_2, "human"), "human", unique_id_2)
  )

pairwise_dist
```

```{r}
# List of emotion columns
emotions <- c("Pride","Anger","Guilt","Shame","Embarrassment","Disgust","Contempt","Compassion")

# Initialize empty tables to store results
all_pairwise <- list()
all_avg <- list()
all_models <- list()

for(em in emotions){
  
  # Step 1: Pairwise differences=============================================
  pairwise_dist <- df_wide %>%
    group_by(stimuli) %>%
    do({
      ids <- .$unique_id
      values <- .[[em]]  
      # Compute absolute pairwise differences
      dist_mat <- outer(values, values, FUN = function(x, y) abs(x - y))
    
      dist_df <- as.data.frame(as.table(dist_mat)) %>%
        rename(unique_id_1 = Var1, unique_id_2 = Var2, distance = Freq) %>%
        filter(unique_id_1 != unique_id_2)
      
      dist_df$unique_id_1 <- ids[as.numeric(dist_df$unique_id_1)]
      dist_df$unique_id_2 <- ids[as.numeric(dist_df$unique_id_2)]
      dist_df
    }) %>%
    ungroup() %>%
    rowwise() %>%
    mutate(pair = paste(sort(c(unique_id_1, unique_id_2)), collapse = "_")) %>%
    distinct(pair, .keep_all = TRUE) %>%
    select(-pair) %>%
    mutate(
      unique_id_1_std = ifelse(startsWith(unique_id_1, "human"), "human", unique_id_1),
      unique_id_2_std = ifelse(startsWith(unique_id_2, "human"), "human", unique_id_2),
      emotion = em
    )
  
  # Step 2: Aggregate by type=============================================
  nonhuman_to_human_avg <- pairwise_dist %>%
    filter((unique_id_1_std == "human" & unique_id_2_std != "human") |
           (unique_id_2_std == "human" & unique_id_1_std != "human")) %>%
    mutate(non_human_id = ifelse(unique_id_1_std == "human", unique_id_2_std, unique_id_1_std)) %>%
    group_by(stimuli, non_human_id) %>%
    summarise(
      unique_id_1 = first(unique_id_1_std),
      unique_id_2 = first(unique_id_2_std),
      distance = mean(distance, na.rm = TRUE),
      emotion = first(emotion),
      .groups = "drop"
    ) %>%
    select(stimuli, unique_id_1, unique_id_2, distance, emotion)
  
  human_to_human_avg <- pairwise_dist %>%
    filter(unique_id_1_std == "human" & unique_id_2_std == "human") %>%
    group_by(stimuli) %>%
    summarise(
      unique_id_1 = first(unique_id_1_std),
      unique_id_2 = first(unique_id_2_std),
      distance = mean(distance, na.rm = TRUE),
      emotion = first(emotion),
      .groups = "drop"
    )
  
  nonhuman_to_nonhuman <- pairwise_dist %>%
    filter(unique_id_1_std != "human" & unique_id_2_std != "human") %>%
    select(stimuli, unique_id_1 = unique_id_1_std, unique_id_2 = unique_id_2_std, distance) %>%
    mutate(emotion = em)
  
  final_pairwise_dist <- bind_rows(nonhuman_to_human_avg, human_to_human_avg, nonhuman_to_nonhuman) %>%
    mutate(pair_id = paste(pmin(unique_id_1, unique_id_2), pmax(unique_id_1, unique_id_2), sep = "_")) %>%
    distinct(stimuli, pair_id, .keep_all = TRUE) %>%
    select(-pair_id)
  
  # Store results
  all_pairwise[[em]] <- final_pairwise_dist
  
  # Step 3: Linear model for human vs non-human=============================================
  df_analysis <- final_pairwise_dist %>%
    mutate(u1 = as.character(unique_id_1),
           u2 = as.character(unique_id_2),
           pair_model = paste(pmin(u1, u2), pmax(u1, u2), sep = "_"),
           pair_type = case_when(
             u1 == "human" & u2 == "human" ~ "human_human",
             u1 == "human" | u2 == "human" ~ "LLM_human",
             TRUE ~ "LLM_LLM"
           )) %>%
    filter(pair_type != "LLM_LLM") %>%
    mutate(pair_type = factor(pair_type, levels = c("human_human", "LLM_human")),
           pair_model = factor(pair_model))
  
  model <- lmer(distance ~ pair_type + (1|stimuli), df_analysis)
  model2 <- lmer(distance ~ pair_model + (1|stimuli), df_analysis)
  
  all_models[[em]] <- list(model=model, model2=model2)
  # Extract coefficients with broom.mixed::tidy()
    coefs1 <- tidy(model) %>% mutate(emotion = em, model_type = "pair_type")
    coefs2 <- tidy(model2) %>% mutate(emotion = em, model_type = "pair_model")
    all_coefs <- bind_rows(all_coefs, coefs1, coefs2)
    
  # Step 4: Average across stimuli=============================================
  avg_pairwise_dist <- final_pairwise_dist %>%
    mutate(u1 = pmin(unique_id_1, unique_id_2),
           u2 = pmax(unique_id_1, unique_id_2)) %>%
    group_by(u1, u2, emotion) %>%
    summarise(avg_distance = mean(distance, na.rm = TRUE), .groups = "drop") %>%
    mutate(u2 = factor(u2, levels = c(sort(setdiff(unique(u2), "human")), "human")))
  
  # Store results
  all_avg[[em]] <- avg_pairwise_dist
}

# Select relevant columns for a clean table =============================================
all_coefs %>%
  mutate(p.value = 2 * (1 - pnorm(abs(statistic)))) %>%
  select(emotion, model_type, term, estimate, std.error, statistic, p.value)%>% 
  nice_table(.)
```

```{r,fig.height=6,fig.width=8}
# Combine all emotions into single long tables
combined_pairwise <- bind_rows(all_pairwise)
combined_avg <- bind_rows(all_avg)
  

ggplot(combined_avg, aes(x = u1, y = u2, fill = avg_distance)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(avg_distance, 2)), size = 3) +
  scale_fill_distiller(palette = "Gn", direction = 1) +
  scale_y_discrete(limits = levels(combined_avg$u2)) +  # DO NOT reverse, keep human first
  labs(
    title = "Average Emotion Differences Across Stimuli",
    x = "", y = "", fill = "Avg Distance"
  ) +
  facet_wrap(~emotion) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# correlation across stimuli 
combined_pairwise <- bind_rows(all_pairwise)
# Pivot to wide format: each row = stimulus, columns = pairs
pair_matrix <- combined_pairwise %>%
  mutate(pair = paste(id_1, id_2, sep = "_")) %>%
  select(stimuli, pair, sim) %>%
  pivot_wider(names_from = pair, values_from = sim, values_fill = list(sim = NA))

# Convert to numeric matrix
stim_matrix <- as.matrix(pair_matrix[,-1])
rownames(stim_matrix) <- pair_matrix$stimuli
# Correlation between rows (stimuli) based on all pairs
cor_matrix <- cor(t(stim_matrix), use = "pairwise.complete.obs")

cor_melt <- melt(cor_matrix)
colnames(cor_melt) <- c("Stimulus1", "Stimulus2", "Correlation")

ggplot(cor_melt, aes(x = Stimulus1, y = Stimulus2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Across Stimuli (All Pairs)",
       x = "Stimulus",
       y = "Stimulus")


```


# alignment analysis 

## correlation by category - does LLM predicted human emotional and behavioral responses to norm violation?

### by gender
### by tie strength
### by moral foundation

## difference in alignment by categories



# item wise analysis

### gender difference (is there a sig difference for female vs. male for each measure (i.e., emotion, action, wrongness, app) - anova)
```{r}

```


### strong vs. weak tie (is there a sig difference for sanction severeness and emotion intensity for strong tie vs. weak tie - anova)

```{r}

```

### if main effect, interactions?
```{r}


```




